---
title: "Module 1: Theory Primer"
subtitle: "Becker and Phelps, with Ride-Sharing Examples"
output:
  xaringan::moon_reader:
    css: [default, default-fonts]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
---

```{css, echo=FALSE}
.remark-code, .remark-inline-code { font-size: 80%; }
.remark-slide-content { padding: 1em 2em; }
.small { font-size: 80%; }
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, fig.align = "center",
  fig.width = 10, fig.height = 4.2, dpi = 150,
  message = FALSE, warning = FALSE
)
library(tidyverse)
theme_set(theme_minimal(base_size = 14))
```

# Course Map

<table>
<tr><th>#</th><th>Module</th><th>Status</th></tr>
<tr><td><b>1</b></td><td><b>Theory Primer: Becker + Phelps</b> <i>(you are here)</i></td><td>← current</td></tr>
<tr><td>2</td><td>Audit & Correspondence Studies</td><td>upcoming</td></tr>
<tr><td>3</td><td>Decomposition Methods (Cook et al. 2021)</td><td>upcoming</td></tr>
<tr><td>4</td><td>Algorithmic Audits</td><td>upcoming</td></tr>
<tr><td>5</td><td>Modern Methods & The Practitioner Question</td><td>upcoming</td></tr>
</table>

---

# The Question

Why does an economic agent treat one group differently from another **even when productivity is identical**?

--

Two classical answers, both still cited 50+ years later:

--

- **Becker (1957):** because the agent has a **taste** against group $B$. Discrimination is preference-driven.

--

- **Phelps (1972):** because the agent has **noisy information** about individuals and rationally falls back on group averages. Discrimination is information-driven.

--

The two answers have **very different policy implications**, and both apply to ride-sharing.

---

# Becker (1957): Taste-Based Discrimination

An employer / coworker / customer dislikes interacting with group $B$ and acts as if hiring (or buying from) a $B$-type costs an extra $d$ — the **discrimination coefficient**.

--

| Source | Who has the taste | Mechanism |
|---|---|---|
| Employer | The hiring manager | Won't hire $B$ even when productive |
| Coworker | $A$-type employees | Demand a wage premium for mixed workplaces |
| Customer | The consumer | Won't buy from a $B$-type |

--

**The famous prediction:** in a competitive market, taste-based discrimination is **eroded by competition**. A non-prejudiced firm hires cheaper $B$-types and undercuts the prejudiced firm.

---

# But Taste-Based Discrimination Persists. Why?

In practice, the competitive-erosion prediction often fails. Several reasons, none fully satisfying:

--

1. **Customer discrimination** — if customers themselves prefer $A$-types, the firm has a *profit* reason to discriminate. Competition makes it **worse**, not better.

--

2. **Coworker discrimination** — workplaces stay segregated because mixing carries a wage premium

--

3. **Search and matching frictions** — markets aren't frictionless, so prejudiced employers don't pay the full cost of their tastes

--

4. **Low-cost discriminatory actions** — if the prejudicial act is cheap (e.g., cancelling a ride), the prejudiced agent doesn't need to be priced out

---

# Becker in the Wild: Ge et al. (2016)

**Field experiment**: matched-pair audit on Uber and Lyft in Boston and Seattle.

--

- Researchers created rider profiles with stereotypically white and stereotypically African-American–sounding names
- Submitted thousands of ride requests
- Measured cancellation rates and wait times

--

**Result:** cancellation rates ~**2× higher** for African-American riders. Wait times longer too.

--

**This is Becker, customer flavor.**

- Drivers are "consumers" of the ride request
- They're acting on a taste-based preference
- Productivity (the fare) is identical across the two profiles
- Competition does **not** erase it: cancelling is essentially free for the driver

--

We'll do this paper properly in Module 2.

---

# Phelps (1972): Statistical Discrimination

Forget tastes. Suppose the employer is *perfectly fair* but faces a **signal-extraction problem**: each candidate has a noisy signal of productivity, and the employer Bayes-updates from group priors.

--

If the priors **differ across groups**, then for the same observed signal, the posterior estimate is **higher for $A$-types than for $B$-types** — and the employer rationally pays them more.

--

**No prejudice. No taste. Just Bayes.**

---

# Phelps: Two Distinct Mechanisms

**1. Different group means.** Group $A$'s productivity distribution is centered higher (for whatever historical reason).

The Bayesian update is the obvious one — for the same signal, the posterior is shifted in $A$'s favor.

--

**2. Different signal precision.** Group $A$ and $B$ have the *same* mean productivity, but the test/résumé/interview is **noisier** for $B$-types.

The Bayesian employer puts more weight on the prior for $B$-types (since the signal is less informative) and less on the noisy individual signal.

--

Productive $B$-types get penalized for having less-informative signals — even though the *group* mean is the same.

---

# Why Phelps Matters More in 2026

**Modern prediction systems are literally statistical-discrimination engines.**

--

- Take a noisy signal (your features)
- Update against a prior estimated from training data
- Output a decision

--

If the training data has different distributions across groups, the model will rationally treat them differently — *exactly as Phelps describes*.

--

The model isn't prejudiced. It's optimizing.

---

# Phelps in the Wild: The Uber Acceptance Model

From Module 2 of the ML course: a race-blind logistic regression for driver acceptance learns that **neighborhood is predictive**.

--

- The model assigns a posterior probability of acceptance to each request
- That posterior depends on the prior estimated from training data
- The prior differs by neighborhood (because base rates of acceptance genuinely differ)
- Therefore the posterior differs systematically by neighborhood — which correlates strongly with race

--

**The model is statistically discriminating. Perfectly. Rationally.** In the sense Phelps had in mind in 1972.

--

The hard question for an applied economist: should we *care* about that, given the model is "just" doing Bayesian updating? Most modern policy frameworks (disparate-impact doctrine) say yes.

---

# Becker vs Phelps: The Cheat Sheet

| | Becker (taste-based) | Phelps (statistical) |
|---|---|---|
| Cause | Prejudice / preference | Rational use of imperfect info |
| Eliminated by competition? | In theory yes; in practice often no | No |
| Disappears with better individual signals? | No | Yes |
| Profit cost to discriminator | Yes | No |
| Empirical fingerprint | Higher returns at non-discriminating firms | Patterns scale with signal noise / prior gaps |

--

**These models are not mutually exclusive.** Both forces operate in any real market. The empirical literature (Module 3) tries to decompose observed gaps into these portions, with limited success.

---

# Coate-Loury (1993): The Self-Fulfilling Equilibrium

A nasty extension of Phelps. If employers **expect** $B$-types to be less productive:

- $B$-types rationally invest **less** in skill (the return is lower)
- This justifies the employers' belief
- The cycle continues

--

Multiple equilibria are possible:
- A "good" one where everyone invests and everyone is paid fairly
- A "bad" one where group $B$ doesn't invest and gets paid less

--

Both are consistent with rational behavior. **Affirmative action** in this model works by *temporarily distorting incentives* to push the system from the bad equilibrium to the good one.

---

# What This Tells the Practitioner

When you're auditing a system, ask:

--

1. Is the discriminator a **person** or an **algorithm**?

--

2. Does the discriminator have a **profit reason** to do it? *(If yes, Becker's competitive-erosion prediction won't work.)*

--

3. Is the **signal** about individuals **noisy**? *(If yes, Phelps says even rational agents will lean on group priors.)*

--

4. Is the **prior** itself a product of past discrimination? *(If yes, you're in Coate-Loury territory and "rational" behavior just perpetuates injustice.)*

--

These four questions are more useful in practice than memorizing formulas.

---
class: inverse, center, middle

# Exercise
### Two Indistinguishable Acceptance Gaps

---

# Setup

We simulate 5,000 ride requests across two groups, $A$ and $B$, where the **true** acceptance probability is identical.

```{r data-gen}
set.seed(2026); n <- 5000
riders <- tibble(
  rider_id  = 1:n,
  group     = sample(c("A","B"), n, replace = TRUE,
                     prob = c(0.6, 0.4)),
  neighborhood = ifelse(
    runif(n) < ifelse(.data$group == "B", 0.65, 0.25),
    "Eastside", "Downtown"),
  hour      = sample(0:23, n, replace = TRUE),
  is_night  = as.integer(hour >= 22 | hour <= 5),
  trip_dist = pmax(rlnorm(n, 1.2, 0.6), 0.3),
  true_p    = plogis(-0.2 - 0.5 * is_night - 0.1 * abs(trip_dist - 4)),
  accepted_baseline = rbinom(n, 1, true_p)
)
```

Note: `group` does **not** appear in `true_p`. The baseline is fair.

---

# Inject Discrimination Two Ways

```{r inject}
# (1) Becker: drivers cancel B requests at +15% regardless of features
riders <- riders |>
  mutate(
    p_becker = ifelse(group == "B", true_p * 0.85, true_p),
    accepted_becker = rbinom(n, 1, p_becker)
  )

# (2) Phelps: race-blind algorithm uses neighborhood as a proxy
fit_phelps <- glm(
  accepted_baseline ~ neighborhood + is_night + trip_dist,
  data = riders, family = binomial
)
riders$p_phelps <- predict(fit_phelps, type = "response")
riders$accepted_phelps <- rbinom(n, 1, riders$p_phelps)
```

---

# Same Observable Gap, Different Mechanism

```{r compare, echo=FALSE}
riders |>
  group_by(group) |>
  summarise(
    Becker = round(mean(accepted_becker), 3),
    Phelps = round(mean(accepted_phelps), 3),
    .groups = "drop"
  ) |>
  knitr::kable()
```

--

A naive audit ("compare acceptance rates by group") would see the same gap in both cases and call both discriminatory.

---

# What an Audit Sees: Marginal vs Controlled

```{r audit, echo=FALSE}
audit_marg_b <- glm(accepted_becker ~ group, data = riders, family = binomial)
audit_marg_p <- glm(accepted_phelps ~ group, data = riders, family = binomial)
audit_ctrl_b <- glm(accepted_becker ~ group + neighborhood + is_night + trip_dist,
                    data = riders, family = binomial)
audit_ctrl_p <- glm(accepted_phelps ~ group + neighborhood + is_night + trip_dist,
                    data = riders, family = binomial)

tibble(
  Mechanism = c("Becker", "Phelps"),
  `Marginal coef on groupB` = round(c(coef(audit_marg_b)["groupB"],
                                       coef(audit_marg_p)["groupB"]), 3),
  `Controlled coef on groupB` = round(c(coef(audit_ctrl_b)["groupB"],
                                         coef(audit_ctrl_p)["groupB"]), 3)
) |> knitr::kable()
```

--

- **Becker**: controlling for features doesn't help — the gap survives
- **Phelps**: controlling for features **eliminates** the gap, because the discrimination is *fully mediated* by neighborhood

This is the empirical problem with Phelps-style proxy discrimination: **regression-based audits miss it.**

---

# Punchline

To detect Phelps-style discrimination you need **either**:

--

- an **experimental audit** (Module 2 — Ge et al. 2016 style), or

--

- a **strong theory** of which features "should" matter (counterfactual reasoning, causal models)

--

Pure regression with controls won't do it. Once the discrimination is laundered through a correlated feature, the regression treats the feature as the explanation.

--

**This is why the empirical literature on discrimination is methodologically fraught.** The next four modules walk through the methods that try to handle it.

---
class: inverse

# The Key Takeaways

<br>

### 1. Becker = preference-driven; Phelps = information-driven. Both happen in the wild.

--

<br>

### 2. Ride-sharing is a clean laboratory for both: Ge et al. for Becker, the acceptance model for Phelps.

--

<br>

### 3. Audits that control for features can miss Phelps. You need a different empirical strategy.

---

# Course Map

<table>
<tr><th>#</th><th>Module</th><th>Status</th></tr>
<tr><td>1</td><td>Theory Primer <i>(just finished)</i></td><td>✓ done</td></tr>
<tr><td><b>2</b></td><td><b>Audit & Correspondence Studies</b></td><td>next</td></tr>
<tr><td>3</td><td>Decomposition Methods</td><td>upcoming</td></tr>
<tr><td>4</td><td>Algorithmic Audits</td><td>upcoming</td></tr>
<tr><td>5</td><td>Modern Methods</td><td>upcoming</td></tr>
</table>

Say **"start module 2"** when ready.
