---
title: "Module 4: Algorithmic Audits"
subtitle: "Pandey-Caliskan, Chen-Mislove-Wilson, and How to Audit a Black-Box Pricing System"
output:
  xaringan::moon_reader:
    css: [default, default-fonts]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
---

```{css, echo=FALSE}
.remark-code, .remark-inline-code { font-size: 80%; }
.remark-slide-content { padding: 1em 2em; }
.small { font-size: 80%; }
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, fig.align = "center",
  fig.width = 10, fig.height = 4.2, dpi = 150,
  message = FALSE, warning = FALSE
)
library(tidyverse)
library(broom)
theme_set(theme_minimal(base_size = 14))
```

# Course Map

<table>
<tr><th>#</th><th>Module</th><th>Status</th></tr>
<tr><td>1</td><td><a href="../module-01/slides.html">Theory Primer</a></td><td>✓ done</td></tr>
<tr><td>2</td><td><a href="../module-02/slides.html">Audit & Correspondence Studies</a></td><td>✓ done</td></tr>
<tr><td>3</td><td><a href="../module-03/slides.html">Decomposition Methods</a></td><td>✓ done</td></tr>
<tr><td><b>4</b></td><td><b>Algorithmic Audits</b> <i>(you are here)</i></td><td>← current</td></tr>
<tr><td>5</td><td>Modern Methods</td><td>upcoming</td></tr>
</table>

---

# A New Kind of Gatekeeper

The audit studies in Module 2 targeted **people**. The decomposition in Module 3 looked at **observational gaps**. This module is about auditing **algorithms**.

--

The algorithm is now the gatekeeper. It sets prices, dispatches drivers, matches riders, suggests pay rates. You can audit it from the outside (scraped data) or the inside (full access + the ability to ship fixes).

--

Two key papers for this module:
- **Pandey & Caliskan (2021)** — disparate impact in Chicago ride-hail pricing
- **Chen, Mislove & Wilson (2015)** — *Peeking Beneath the Hood of Uber*, the methodological template

---

# What's Different About Algorithmic Audits

**1. The "manipulation" is querying the system, not deceiving a person.** No fraud, no IRB headache.

--

**2. You can run *millions* of queries.** Pandey-Caliskan analyze ~100M Chicago trips.

--

**3. The "matched pair" is a counterfactual query.** "What would the algorithm have done if everything were the same except this one feature?"

--

**The catch:** you don't usually *control* the inputs. You observe what real users get. So you're back in regression-and-controls territory, with all the bad-controls problems that implies.

---

# Pandey & Caliskan (2021)

**Title:** *Disparate Impact of AI Bias in Ridehailing Economy's Price Discrimination Algorithms*

**Setting:** Chicago. The city publishes anonymized trip data: fare, distance, duration, pickup/dropoff census tracts.

--

**Audit question:** *Are pickup or dropoff areas with higher minority populations charged more per mile for equivalent trips?*

--

**Method (shadow audit):**
1. Compute fare-per-mile per trip
2. Merge trips with census-tract demographics
3. Regress fare-per-mile on demographic features, controlling for trip observables

--

**Result:** Trips with pickup or dropoff in higher-minority census tracts have **higher fare-per-mile**, even after controls.

---

# The Implicit Mechanism

Uber's surge-pricing algorithm uses real-time supply/demand signals. **The algorithm never reads "race."** But:

--

- In neighborhoods with thinner driver supply, prices surge more often

--

- Driver supply is correlated with neighborhood demographics (historical patterns + current urban form)

--

- So the surge signal is correlated with race even though the algorithm is race-blind

--

This is **statistical discrimination by an algorithm**, mediated by a proxy. **Phelps's model running in real time.**

---

# What Pandey-Caliskan Identifies (and Doesn't)

| What | Yes? |
|---|---|
| Correlation between demographics and price-per-mile | **Yes** |
| Documented disparate impact | **Yes** |
| Causal mechanism | **No** |
| Policy fix | **No** |

--

The paper produces a quantitative claim that can be debated by regulators. That's its job. It is not the final word on whether the algorithm is "discriminatory" — that's a normative call.

---

# Chen, Mislove & Wilson (2015)

**Title:** *Peeking Beneath the Hood of Uber* (IMC '15)

The original surge-pricing audit. Not about discrimination per se, but the **methodology** is the template for everything that followed.

--

**The setup:**
- Built a measurement infrastructure around Uber's app
- Polled surge multipliers and driver counts at hundreds of locations every few seconds, for weeks
- Recorded the spatial structure of surge

--

**Findings:**
- Surge cells are tiny (~0.5 km × 0.5 km)
- Surge changes rapidly (5–10 min durations)
- Surge increases supply only marginally — most rebalancing comes from rider cancellations

--

**Methodological contribution:** *you can audit a black-box pricing system using only the public-facing API*. Subsequent algorithmic audits have all built on this template.

---

# How to Audit a Pricing Algorithm: 4 Steps

**Step 1.** Define the disparate-impact question precisely. *Whose price? Per what? Conditional on what? Compared to what?*

--

**Step 2.** Pick the data source. Internal (trip log + dispatch log + linked Census) or external (scraped surge / public city data).

--

**Step 3.** Pick the comparison.
- Within-trip counterfactual ("if this trip were in a different neighborhood")
- Within-rider FE
- Cross-sectional regression (Pandey-Caliskan style)

--

**Step 4.** Report carefully. Point estimate + SE + robustness + power + interpretation. Honest reports always include all five.

---

# External vs Internal Audits

| | External | Internal |
|---|---|---|
| Sample | Small, biased | Universe of decisions |
| Variables | Public observables | Everything the system uses |
| Counterfactuals | Hard (need a model) | Easy (run an A/B test) |
| Reportability | Free to publish | Constrained |
| Can ship fixes? | No | Yes |

--

The interesting move for an **internal economist**: bring the external auditor's questions in-house and run them on the full data with the ability to ship the fix.

---
class: inverse, center, middle

# Exercise
### Stylized Surge-Pricing Audit

---

# Setup: 30 Neighborhoods, Different Demographics

```{r data-gen}
set.seed(2026)
n_nbhd <- 30
city <- tibble(
  neighborhood  = paste0("N", sprintf("%02d", 1:n_nbhd)),
  pct_minority  = runif(n_nbhd, 0.05, 0.90),
  driver_supply = pmax(8 - 6 * pct_minority + rnorm(n_nbhd, 0, 0.5), 0.5),
  demand        = rnorm(n_nbhd, 10, 1.5)
) |>
  mutate(
    surge_multiplier = pmax(1, demand / driver_supply),
    fare_per_mile    = 1.50 * surge_multiplier
  )
```

The pricing rule is `pmax(1, demand/supply)`. **Race is not in this formula.** But supply correlates with race because of how the city is set up.

---

# The Audit Regression

```{r audit-reg}
audit_fit <- lm(fare_per_mile ~ pct_minority, data = city)
```

```{r audit-tbl, echo=FALSE}
tidy(audit_fit) |>
  mutate(across(where(is.numeric), \(x) round(x, 4))) |>
  knitr::kable()
```

--

Significant positive coefficient on `pct_minority` despite the algorithm never seeing race — the disparate impact is mediated entirely through the supply channel.

---

# The Bad-Controls Trap (Again)

```{r bad-control}
audit_ctrl <- lm(fare_per_mile ~ pct_minority + driver_supply, data = city)
```

```{r bad-tbl, echo=FALSE}
tidy(audit_ctrl) |>
  mutate(across(where(is.numeric), \(x) round(x, 4))) |>
  knitr::kable()
```

--

Adding `driver_supply` as a control kills the demographic coefficient — *because driver_supply is the mechanism*. The controlled regression hides the disparate impact instead of explaining it.

---

# Two Policy Fixes

```{r fixes, echo=FALSE}
city <- city |>
  mutate(
    fare_no_surge = 1.50,
    fare_capped   = 1.50 * pmin(surge_multiplier, 1.5),
    group         = ifelse(pct_minority >= median(pct_minority), "high", "low")
  )

di <- function(col) {
  grouped <- city |> group_by(group) |>
    summarise(avg = mean({{col}}), .groups = "drop")
  round(grouped$avg[grouped$group == "high"] /
        grouped$avg[grouped$group == "low"], 3)
}
revenue <- function(col) round(sum(city[[col]] * city$demand), 1)

tibble(
  Policy = c("Surge as-is", "Cap surge at 1.5×", "No surge at all"),
  `DI ratio (high/low)` = c(di(fare_per_mile), di(fare_capped), di(fare_no_surge)),
  `Total revenue` = c(revenue("fare_per_mile"),
                      revenue("fare_capped"),
                      revenue("fare_no_surge"))
) |> knitr::kable()
```

--

Closing the disparate-impact gap costs revenue. The math draws the curve; **humans pick the point**.

---
class: inverse

# The Key Takeaways

<br>

### 1. Algorithmic audits can be run at scale, on real data, with low ethical cost — they are the dominant modern methodology.

--

<br>

### 2. Pandey-Caliskan documents disparate impact in ride-hail pricing without claiming a causal mechanism.

--

<br>

### 3. The bad-controls problem follows you into the algorithmic world: regressions with mediator controls hide the very mechanism you're trying to find.

---

# Course Map

<table>
<tr><th>#</th><th>Module</th><th>Status</th></tr>
<tr><td>1</td><td><a href="../module-01/slides.html">Theory Primer</a></td><td>✓ done</td></tr>
<tr><td>2</td><td><a href="../module-02/slides.html">Audit & Correspondence Studies</a></td><td>✓ done</td></tr>
<tr><td>3</td><td><a href="../module-03/slides.html">Decomposition Methods</a></td><td>✓ done</td></tr>
<tr><td>4</td><td>Algorithmic Audits <i>(just finished)</i></td><td>✓ done</td></tr>
<tr><td><b>5</b></td><td><b>Modern Methods & Practitioner</b></td><td>next</td></tr>
</table>

Say **"start module 5"** when ready.
